Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-03-06 draft.tex
Added Parts of Data Chapter

Longitudinal population-based studies like the Study of Health in Pomerania \cite{Volzke2011} aim to gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
The sample is drawn randomized to avoid selection bias prohibit statements based on statistical correlations in the cohort to be inferred to the whole population.
%
Also a information bias needs to be avoided by strictly standardizing the data acquisition.
%
Statistical correlations are also prone to confounding, meaning that two factors influence each other and therefore should be normalized with respect to each other.
%
When for example one investigates risk factors for prostate cancer in male subjects the outcome is strongly dependent on the age.
%
Therefore results need to be age adjusted to be comparable.
%
Confounding variables, however, are often not obvious at all and characterizing them is already an epidemiological result.
\\\\
Modern cohort studies often include medical image data which introduces new problems.
%
Since it is unethical to expose people to radiation, non-harming imaging like Magnetic Resonance Imaging (MRI) or Ultrasound Imaging is used.
%
As MRI scans are expensive there exists a tradeoff between quality of the image data and their associated costs.
%
To quantify these data it is necessary to segment it.
%
Manual segmentation via radiological experts is possible but very costly and prone to inter- and intra observer variability.
%
Segmentation algorithms allow for (semi)-automated analysis of the data but require sophisticated methods due to high inter-subject variability caused by the subject diversity.
%
Analyzing spatial data with other epidemiological factors require techniques which reach beyond standard statistical methods.
\\
We propose a Interactive Visual Analysis approach to provide a way to analyze both image- and non-image data.
%
Visual queries and direct feedback of Visual Analytics systems allow for a fast exploration of the data space.
%
Intended as a extension to the well established epidemiological tools it provides a way to rapidly validate hypothesis as well as trigger hypothesis generation using Data Mining methods such as clustering.
%

@@

Reproducibility of results is an epidemiological key requirement.
%
Longitudinal epidemiological studies require the acquired attributes to be comparable to evaluate them.
%
If the data acquisition process changes, a information bias is introduced to the data, disallowing inference between acquisition cycles.
%
This underlines the high quality standards to methods processing the data, whether to extract additional parameters or gain insight.
%
To determine, whether a subject is prone to be affected by a certain disease, relative risks a expressed through the evaluation of p values which indicate statistical significance.
%
Statistics tools such as SPSS and STATA play a major role for analyzing epidemiolgical data. 
%
Graphic data representation is largely used to show results rather than gaining insight.

Bernhards Paper\\
- Missing Data\\
- Follow-Up Questions?\\
- grouping essentiell\\
- problems when analyzing image data\\
	- what are Problems there\\
	- how can shape be included?\\
	- show image data of multiple subjects

\subsection{Epidemiological Data}
Benrhard Paper
- sociodemographic data
- medical data
- pain indicators
- DATA TYPES
	- dichotomous data

>>>>

Longitudinal population-based studies like the Study of Health in Pomerania \cite{Volzke2011} aim to gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
The sample is drawn randomized to avoid selection bias prohibit statements based on statistical correlations in the cohort to be inferred to the whole population.
%
Also a information bias needs to be avoided by strictly standardizing the data acquisition.
%
Statistical correlations are also prone to confounding, meaning that two factors influence each other and therefore should be normalized with respect to each other.
%
When for example one investigates risk factors for prostate cancer in male subjects the outcome is strongly dependent on the age.
%
Therefore results need to be age adjusted to be comparable.
%
Confounding variables, however, are often not obvious at all and characterizing them is already an epidemiological result.
\\\\
Modern cohort studies often include medical image data which introduces new problems.
%
Since it is unethical to expose people to radiation, non-harming imaging like Magnetic Resonance Imaging (MRI) or Ultrasound Imaging is used.
%
As MRI scans are expensive there exists a tradeoff between quality of the image data and their associated costs.
%
To quantify these data it is necessary to segment it.
%
Manual segmentation via radiological experts is possible but very costly and prone to inter- and intra observer variability.
%
Segmentation algorithms allow for (semi)-automated analysis of the data but require sophisticated methods due to high inter-subject variability caused by the subject diversity.
%
Analyzing spatial data with other epidemiological factors require techniques which reach beyond standard statistical methods.
\\
We propose a Interactive Visual Analysis approach to provide a way to analyze both image- and non-image data.
%
Visual queries and direct feedback of Visual Analytics systems allow for a fast exploration of the data space.
%
Intended as a extension to the well established epidemiological tools it provides a way to rapidly validate hypothesis as well as trigger hypothesis generation using Data Mining methods such as clustering.
%
\textbf{Characterization of the healthy aging process! Which change indicates a unusual pathological change?}

@@

Reproducibility of results is an epidemiological key requirement.
%
Longitudinal epidemiological studies require the acquired attributes to be comparable to evaluate them.
%
If the data acquisition process changes, a information bias is introduced to the data, disallowing inference between acquisition cycles.
%
This underlines the high quality standards to methods processing the data, whether to extract additional parameters or gain insight.
%
To determine, whether a subject is prone to be affected by a certain disease, relative risks a expressed through the evaluation of p values which indicate statistical significance.
%
Statistics tools such as SPSS and STATA play a major role for analyzing epidemiolgical data. 
%
Graphic data representation is largely used to show results rather than gaining insight.
\\
Group subjects using epidemiological factors is essential in order to make statements about their statitical power.
%
Grouping is carried out hypothesis driven.
%
Age for example is also divided into groups (\emph{binned}) when investigating its influence on a condition.
%
These groups depend strongly on the condition of interest and therefore there is no defined standard on how to categorize these values.

\subsection{Epidemiological Data}
Epidemiological data is highly heterogenous.
%
Information about medical history and examinations, genetic conditions, geographical data, questionaire answers, and image data yield complex data spaces for each subject.
%
Often data are derived from acquired variables to either group or threshold values or get information derived from reviewed data such as breast density data for women.
%
This underlines also the problem of missing data since for ethical, legal or medical reasons some data variables can not be gathered for each subject.
%
Follow-up examinations or -questions for conditions also produce variables only available for a small amount of subjects.
%
\\
Indicators for medical conditions as well as questions about a subjects lifestyle are also often \emph{dichotomous}, meaning that they only have two manifestations (often \emph{Yes} or \emph{No}).
%
This allows for the calculation of \emph{odds ratios} which describe the relation of two \emph{dichotomous} variables, allowing for direct comparison of their influences.
%
Dichotomous data can also be derived by combining aggregating data variables to yield only two manifestations (e.g. subjects younger or older than 50).

Bernhard Paper\\
- sociodemographic data\\
- medical data\\
- pain indicators\\
- DATA TYPES\\
	- dichotomous data\\
\\
- Missing Data\\
- Follow-Up Questions?\\
- grouping essential\\
- problems when analyzing image data\\
	- what are Problems there\\
	- how can shape be included?\\
	- show image data of multiple subjects
